---
title: "Week 1 R Challenge Answer"
date: "2024-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Challenge:

Write a function that uses a Riemann Sum to correctly estimate $\pi$ to at least 5 correct decimals: $3.14159$

Ideally, you are combining your knowledge of algebra, geometry, and R loops to estimate $\pi$.

You may *not* use any packages outside of base R, or any trigonometric functions or other pre-existing R functions that have $\pi$ *baked* into them. 

## My Solution:

We know that $x^2 + y^2 = 1$ forms a circle of radius $1$ and area $\pi$. 

Estimate the area between $0$ and $1$ under the curve of $y = \sqrt{1 - x^2}$ using a Riemann Sum. Call this area $A$. Therefore we estimate $\pi$ as 

$$\pi \approx 4 A$$



```{r code}

# Function for y = \sqrt{1 - x^2}
quart_circ = function(x) {
  
  y = sqrt(1 - x^2)
  
  return(y)
}


# Riemann Sum Estimator
riemann_sum = function(n) {
  # n tells us the number of steps to take
  
  steps = seq(from = 0, to = 1, by = 1/n)
  
  area_data = rep(NA, n)
  
  for (i in 1:n) {
    # what is y for a given x?
    y = quart_circ(steps[i])
    
    # multiply to get area of rectangle, and save area
    area_data[i] = y*(1/n)
  }
  
  pi = 4*sum(area_data)
  
  return(pi)
}



riemann_sum(1000000)

```





