---
title: "Lab 8: Maps and Spatial Data"
date: "2024-05-31"
output: pdf_document
---


# Creating Maps in R


The [maps package](https://cloud.r-project.org/web/packages/maps/maps.pdf)


```{r packages}

# load packages
library(pacman)
p_load(maps)

```


## U.S. City Data

The `us.cities` dataset in the `maps` package contains information on the location of about 1000 cities in the U.S. along with information on population and whether the city is a state capital

```{r us city}

head(us.cities)

# Select the subset of cities that are state capitals
capitals = subset(us.cities, capital == 2)


# Start a map with the map() function
map(database = "usa") 
  # to apply something to this map, start a function in the next line
points(x = capitals$long, y= capitals$lat, col = "blue")
  # and another
title("US state capitals")
  # it does this until there's a function that doesn't apply


```


We can *add* a map on top of another map using `add = T` in another `map()` function below the first one (note that `add = F` is the default). For example, let's overlap the U.S. capitals map on top of the world map:

```{r world us cities}

map(database = "world")
map(database = "usa", add = T) 
  # to apply something to this map, start a function in the next line
points(x = capitals$long, y= capitals$lat, col = "blue")
title("US state capitals in the World")

```

You may notice that this map includes Hawaii and Alaska, when the previous map did not. 

## Other Datasets in the `maps` Package 

### The WORLD

```{r world}

# Data of cities around the world
head(world.cities)


# 
map(database = "world")
points(x = world.cities$long, y= world.cities$lat, col = "blue")
title("Map of Cities in maps Package")


```

We can change the look of the dots within the `points()` function:

* `pch = 1, 2, ...` The difference numbers determine the shape. Lots of strange shapes that make you wonder "Who would ever use this and why?" You can find the options [here](http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r)

* `cex = 1` Continuous variable that determines the size of the city point on the map


Let's recreate that map from above but make the points filled in circles (`pch = 20`) and make the points much smaller (`cex = 0.1`)

```{r nicer world}

map(database = "world")
points(x = world.cities$long, y= world.cities$lat, col = "blue",
       pch = 20, cex = 0.1) 
  # also try cex = 0.01
  # add scale
map.scale()

```

### China

China has specific labels for cities where: 0 for non-capital, 1 for capital, 2 for China Municipalities, and 3 for China Provincial capitals

```{r China maps 1}

# Tell it the scope of your map
map("world", "china")
  # non capital cities
map.cities(country = "China", capitals = 0, 
      pch=20, col="blue")
  # Provincial capitals
map.cities(country = "China", capitals = 3, 
      pch=20, col="green")
  # Municipal Capitals
map.cities(country = "China", capitals = 2, 
      pch=20, col="red")
  # Country capital
map.cities(country = "China", capitals = 1, 
      pch=20, col="orange")

```


### State Data

Back to the U.S. 

```{r states 1} 

# We can look at all of the (lower 48, sorry Hawaii and Alaska) states 
map(database = "state")


# We can get specific regions:
map(database = "state", regions= "Oregon")

map(database = "state", regions= c("Oregon", "Washington", "California"))


# Let's plot the most populous city in each of these 3 states:
  # first:
p_load(dplyr)

  # second:
west_city = us.cities %>% 
    # filter to the states we want
  filter(country.etc %in% c("OR", "WA", "CA")) %>%
    # group by each state to find the most populous
  group_by(country.etc) %>%
    # get the most populous city in each state
  slice_max(order_by = pop, n = 1) %>% 
      # change n = 1 to 2 or 3 to get the top 2 or 3 most populous
  ungroup()

  # Now plot:
map(database = "state", regions= c("Oregon", "Washington", "California"))
points(x = west_city$long, y = west_city$lat, 
       col = "blue", pch = 20)



```


### Also the `usmap` Package

The `usmap` package gives you more range of options (including county maps) and works in tandum with `ggplot2`, a package we're familiar with.

```{r usmap counties}

# Load packages
p_load(usmap, ggplot2)

# Plain map
plot_usmap(regions = "counties")

# Add some ggplot elements
plot_usmap(regions = "counties") +
  labs(title = "US Counties") +
  theme_bw()

```

Kind of misleading with Hawaii and Alaska in relation to the longitude and latitude labels, but it's cool. 



# Showing Data

Now let's incorporate economic data into these maps! 

Go to the Github website and scroll down to Week 9 and click on "Spatial Data" and then "raw" and copy the URL. This is inflation data about all US states that comes from [here](https://www.jec.senate.gov/public/index.cfm/republicans/state-inflation-tracker)



```{r get data}
# recall from webscraping lab
p_load(readr)

# URL
urlfile = "https://raw.githubusercontent.com/ojetton/Econometrics_Lab_Spring_24/main/state_inflation_data"

# the data
inflation = read_csv(url(urlfile))

```

We are going to select the columns labeled "State Code" and "Increase in Prices (%) since January 2021," and then merge with the "state" data from the packages we used above. Then we will make a *heat map* of the states where a darker red signifies higher inflation.

```{r inflation map}


inflation = inflation %>% 
    # select the 2nd and 4th columns
              select(1,4) %>%
    # rename columns
              rename_all(~c("state", "inflation")) %>%
    # make state names lowercase
              mutate(state = tolower(state))

# get spatial data about states
state_data = map_data("state")

# Merge datasets
  # note the different names for states in the datasets
state_map = merge(x = inflation, y = state_data, by.x = "state", by.y = "region")



# Plot in ggplot:
ggplot(state_map, aes(x = long, y = lat, group = group, fill = inflation)) +
  geom_polygon(color = "black") +
  scale_fill_continuous(low = "white", high = "red", name = "Inflation % since Jan. 2021") + 
  theme_void()

```



