---
title: "Week 1 Lab"
date: "2024-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown

## What is R Markdown?

R Markdown is a format for writing reports and presentations that include R Code. It can create files in html, pdf, Word, slideshows, and other formats. 


## Writing in R Markdown

Unlike R script files, writing on a plain line in R Markdown will appear as wording in the final file. Here are a few options for word stylization:

* *Italics*: Single asterisks on either side of the word/phrase: $*\text{word}*$

* **Bold**: Double asterisks on either side of the word/phrase: $**\text{word}**$

* Headings: Start a line with a hashtag/pound sign. More hashtags = smaller heading: $\#\text{heading}$

# Heading

## Heading

### Heading

#### Heading

* Lists: Single asterisk, number, or other sign: $*, +, -, 1$


## LaTeX in R Markdown

If you want even more options for writing, such as writing math equations, in R Markdown, it uses \LaTeX 

Many of the writing formatting operations can be performed with LaTeX

\begin{itemize}
  \item Such as writing lists
  \item Or \emph{italicizing} which can be done with $\backslash \text{emph{word}} ~ \Rightarrow $ \emph{word}
  \item Or \textbf{bolding} which can be done with $\backslash \text{textbf{word}} ~ \Rightarrow $ \textbf{word}
\end{itemize}

Most importantly, it can be used for writing math equations. In line with \textdollar e = m c ^ 2 \textdollar $\Rightarrow e = mc^2$ 

Or as its own centered body of text: \textdollar $\backslash$sum_$\text{{t=0}}$^$\backslash$infty $\backslash$beta^t u(c_t) \textdollar $\Rightarrow$
$$\sum_{t=0}^\infty \beta^t u(c_t)$$

Greek letters can also be displayed through LaTeX's math mode. Lowercase first letter yields lowercase greek letter, uppercase first letter yields upper case greek letter:  $\backslash$omega $\Rightarrow \omega$, $\backslash$Omega $\Rightarrow \Omega$ 



## Coding Terminology

When creating **chunks** of code, you have the following options, which are set to **T** by default.

* eval: (T/F) run/evaluate the code and display results

* echo: (T/F) display code and results (runs code either way)

* warning: (T/F) display warnings

* error: (T/F) display errors


## Knitting

The **Knit** button generates the document (in the selected format) with the writing and embedded code chunks and their results. 

Knitting to html is automatically built into R Markdown. To knit to pdf, you will need to download \LaTeX onto your computer (which is a large file and takes time)


## Example
 
Let's do an example file and practice:

* formatting the file

* creating code chunks

* running code in R Markdown

* displaying plots

* knitting


```{r title}
# Load car data
car_data = cars

# And ggplot2
library(ggplot2)

```

```{r eval plot}

ggplot(cars, aes(x = speed, y = dist)) +
  geom_point()

```

Now plot with eval = F

```{r dont eval plot, eval = F}

ggplot(cars, aes(x = speed, y = dist)) +
  geom_point()

```

No plot?


## Review Functions

We will briefly review creating functions in R, since they will come in handy during this term, and some problem sets will specifically ask for functions that perform a certain task.

#### Basics:

```{r function 1}

# function name = function(input) {
#                 operations on input = output 
#                 return(output)
# }

```

#### Example:

Create a function of this function:

$$f(x) = \frac{x - 5}{4}$$
```{r function 2}

# Create the function
f_x = function(x) {
  
  ans = (1/4)*(x - 5)
  
  return(ans)
}


# Try it on x = 10
f_x(10)
```

Now suppose we want this new function:

$$g(x,y) = f(x) - 3y$$

You could do this by plugging in $f(x) = \frac{x - 5}{4}$ is, OR a faster way is to put our previous function $f\_x$ into this new function!

```{r function 3}

# Create function
g_xy = function(x, y) { # note: two inputs
  
  ans = f_x(x) - 3*y
  
  return(ans)
}

# Try with x = 10, y = 2
g_xy(10, 2)
```


## For Loops

We briefly discussed *for* loops last term, but didn't use them. You will use them this term, so let's review.

#### Basics

```{r for 1}

# for (i in start:end) {
#   operation performed on each value of i
# }

```

That's rather abstract so it's easier to start with an example. Apply our function $f\_x$ to every integer from 1 to 20 and print the answer.

```{r for 2}

for (i in 1:20){
  print(f_x(i))
}

```

#### Handy functions for *for* loops

We typically don't want to just perform operations on integers. We can get any linear sequence of numbers we want using the $seq()$ function

Suppose we want every hundredth place from 0 to 1: 0, 0.01, 0.02, ... , 0.98, 0.99, 1

```{r seq}

# Basically
##  seq(from = start, to = end, by = step)

seq(from = 0, to = 1, by = 0.01)

```

We also often want a vector to put our output into. The function $rep(value, n)$ repeats whatever value you give it, n times

```{r rep}

# "empty" vector
rep(NA, 10)

```

In a *for* loop, we can find the $i^{\text{th}}$ value of a vector by using square brackets

For example: apply $f\_x$ to each value in the sequence above using a *for* loop, and store the answers into a new vector

```{r for 3}

seq_1 = seq(from = 0, to = 1, by = 0.01)

seq_2 = rep(NA, 101)

for (i in 1:101){
  seq_2[i] = f_x(seq_1[i])
}

seq_2
```



#### Combine functions and for loops

Let's create a function that creates the first N numbers of the Fibonacci sequence, defined as:

$$F_1 = 0, ~ F_2 = 1, ~ F_{n} = F_{n-1} + F_{n-2}$$

```{r fibonacci}

# function
fibonacci = function(N) {
  
  data = c(0, 1, rep(NA, N - 2))
  
  for (i in 3:N) {
    data[i] = data[i - 1] + data[i - 2]
  }
  
  return(data)
}

# suppose we want the first 20 values (including 0 and 1)
fibonacci(20)

```


