---
title: 'Lab 8: Tables in R'
date: "2024-05-24"
output:
  html_document: default
  pdf_document: default
---

We have talked previously about summarizing data into aesthetic visuals, or graphs. But sometimes, what is important cannot be transformed into a pleasing picture, and requires a *table.*

Today we will discuss how to create tables that succinctly convey the desired information. 

# Tables

We will focus on the following types of tables today:

1. Summary tables: show summary statistics of data

2. Regression tables: report the results of one or multiple regressions


## Summary Tables

### Frequency Statistics

```{r data}

# Use the fivethirtyeight package
library(pacman)
p_load(fivethirtyeight, tidyverse) # tidyverse for cleaning

# Data on Congress members from the 113th congress (2013 - 2015)
age_data = congress_age %>% filter(congress == 113)

table(age_data$party, age_data$chamber)

```

This is nice! But it can look better. Let's make a table reporting the average age for each party within each chamber of Congress.

```{r age table}
# new package: gt
p_load(gt)

age_table = age_data %>%
  group_by(chamber, party) %>%
  summarize(average_age = mean(age)) %>%
  pivot_wider(
    names_from = chamber,
    values_from = average_age
  )

age_table

# make a gt object
age_table = age_table %>% gt()

age_table

```

Much better! We can make more changes:

* Add a title

* Capitalize the first letter of each word

* Write out the full name of each party (D $\rightarrow$ Democrat, etc.)

* Change the number of decimals to 2 for each entry


```{r age table 2}

# from the gt packages
age_table %>%
  
  # title
  tab_header( title = "Average Age of Congress Members by Party and Chamber",
    subtitle = "In the 133th Congress") %>%
  
  # capitalize first word
  cols_label(party = "Party", house = "House", senate = "Senate") %>%
  
  # full party names
  text_case_match("D" ~ "Democrat", "R" ~ "Republican", "I" ~ "Independent") %>%
  
  # decimals
  fmt_number(decimals = 2)
  
  
```



## Regression Tables

One of the most common types of tables you'll have in presentations, reports, or research papers are those reporting the results of a regression (or multiple regressions). 

For this example, we will use the `candy_rankings` dataset also from the `fivethirtyeight` package. It contains survey data from over 250,000 random match-ups between candies, the percent that each candy "won," and information on each candy (including if it contains chocolate, comes in a bar form, is fruity, etc.)

Suppose we want to know if having chocolate or peanuts/almonds makes a candy more likely to be chosen, or if the combination of chocolate and peanuts is more important. We can do this with two regressions:

$$winpercent \sim chocolate + peanutyalmondy$$

$$winpercent \sim chocolate + peanutyalmondy + chocolate \times peanutyalmondy$$

```{r candy regs}
candy = candy_rankings

reg_1 = lm(data = candy, formula = winpercent ~ chocolate + peanutyalmondy)

reg_2 = lm(data = candy, 
           formula = winpercent ~ chocolate + peanutyalmondy + chocolate:peanutyalmondy)


```


#### tab_model() function

This produces an HTML output, so it won't look pretty if you knit to a PDF. However, it looks good when using it in an R script file or knitting to HTML.

```{r tabmodel}
# From package: sjPlot
p_load(sjPlot)

# Basic table
tab_model(reg_1, reg_2)

```


Multiple adjustments can be made

* show.se = (TRUE / FALSE, whether to show standard errors of estimates)

* show.stat = (TRUE / FALSE, whether to show t-statistic)

* show.ci = (TRUE / FALSE, whether to show confidence interval)

* show.p = (TRUE / FALSE, whether to show p-value)

* pred.labels = c("list names of predictor variable(s)")

* dv.labels = c("list names of *dependent* variable(s)")

* string.pred = "Name for *Predictors* column"

* string.ci = "Name for *CI* column"

* string.p = "Name for *p* column"



Let's use some of these options to make our table prettier

1. Show estimates, standard errors, and p-value

2. Rename variables to "Chocolate," "Nuts," and "Chocolate and Nuts." Also change the dependent variable to "Win Percent"

3. Rename columns appropriately

```{r tabmodel 2}

tab_model(reg_1, reg_2,
        # automatically reports estimate, CI, p)
          show.ci = F,
          show.se = T,
        # rename variables
          pred.labels = c("Intercept", "Chocolate", "Nuts", "Chocolate and Nuts"),
          dv.labels = c("Win Percent", "Win Percent"), # wants it twice bc 2 models
        # Rename predictors column
          string.pred = "Variables",
          string.se = "Std. Err."
)

```


#### stargazer package

Also produces multiple types of output: text, html

```{r stargazer}

p_load(stargazer)

stargazer(reg_1, reg_2, type='text')

```


To get nice regression tables in an Rmarkdown file, include `results = 'asis'` in the brackets at the start of the chunk

```{r stargazer 2, results='asis'}
# include results='asis' in {}

# type = html
stargazer(reg_1, reg_2, type='html')

```


There are a lot of options for changing the table within the stargazer package.

```{r stargazer 3, results='asis'}

stargazer(reg_1, reg_2, type='html',
          # add a title
          title = "Candy Regressions",
          
          # label variables (covariates)
          covariate.labels = c("Chocolate", "Nuts", "Chocolate and Nuts", "Intercept"),
          
          dep.var.labels = "Win Percent",
          
          # number of digits in each number
          digits = 2,
          
          # choose which statistics to keep (n and r-squared)
          keep.stat = c("n","rsq"),
          
          ci.level=0.90
          )

```




Can also set type = 'latex' to get the table in latex. Speaking of which....

```{r stargazer 4}

stargazer(reg_1, reg_2, type='latex')

```



## Tables in LaTeX

There may be cases in the future where you need a table for something unrelated to data, or it is showing qualitative information that can't be easily turned into a data frame and then into the table. Luckily, LaTeX has us covered! We can make all sorts of tables. Here's how the tables work:

* Start by centering your table with $\backslash \text{begin \{ center \}}$ and at the end: $\backslash \text{end \{ center \}}$

* Create the table: $\backslash \text{begin \{ tabular \} \{  \}}$

    - In the second set of brackets, use a $c$ to denote an individual column, and a $|$ (pipe symbol?) to denote a vertical line in the table.

* Use $\backslash \text{hline}$ to create a horizontal line 

* Use ampersands $\&$ to *align* the lines of your columns

* End each line with $\backslash \backslash$ to let it know that's the end of that line






